Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-31 manuscript/HIAs.qmd
include ditching

```{r getRelevantNTs}
#| eval: false

@@

# Getting the nature types that have 7FA
myVars <- "7FA"

# naturetypes_summary2 <- naturetypes_summary %>%
#  rowwise() %>%
#  mutate(keepers = sum(c_across(
#    all_of(myVars))>0, na.rm=T)) %>%
#  filter(keepers >0,
#         Ecosystem == "våtmark") %>%
#  select(Nature_type, NiN_mainType, Year)
#
# naturetypes_summary2$Nature_type

# Getting the nature types that have the ADSV variables
myVars2 <- c("7TK", "7SE", "PRTK", "PRSL")
# naturetypes_summary3 <- naturetypes_summary %>%
#  rowwise() %>%
#  mutate(keepers = sum(c_across(
#    all_of(myVars2))>0, na.rm=T)) %>%
#  filter(keepers >0,
#         Ecosystem == "våtmark") %>%
#  select(Nature_type, NiN_mainType, Year)
#
# naturetypes_summary3$Nature_type
```

```{r relevantNTs}
#| eval: false

# These are the nature types I'm keeping for 7FA.
# Note the absence of ombrotriphic types
alienNTS <- c(
  "Slåttemyr",
  "Semi-naturlig våteng",
  "Rik åpen jordvannsmyr i mellomboreal sone",
  "Rik åpen sørlig jordvannsmyr",
  "Rik åpen jordvannsmyr i nordboreal og lavalpin sone"
)

# And these are the nature types that we use for the ADSV indicator
advsNTs <- c( # fens
  "Slåttemyr",
  "Sørlig slåttemyr",
  "Rik åpen jordvannsmyr i mellomboreal sone",
  "Rik åpen sørlig jordvannsmyr",
  "Semi-naturlig myr",
  "Rik åpen jordvannsmyr i nordboreal og lavalpin sone",
  # "Sørlig kaldkilde", # excluded springs

  # bogs (mainly)
  "Høgereligende og nordlig nedbørsmyr",
  "Øyblandingsmyr",
  "Konsentrisk høymyr",
  "Sørlig nedbørsmyr",
  "Atlantisk høymyr",
  "Terrengdekkende myr",
  "Eksentrisk høymyr",
  "Kanthøymyr",
  "Platåhøymyr",
  "Palsmyr"
)
```

@@

# Impart polygon data set
naturetypes <- sf::st_read(dsn = path_naturetypes)
# 117k polygons (2023)
# sometimes this functin struggles with Norwegian letters.
# Restart R and try again.

naturetypes <- naturetypes |>
  # keep only wetlands and our selected nature types
  filter(
    hovedøkosystem == "våtmark",
    naturtype %in% c(advsNTs, alienNTS)
  ) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep = ",") |>
  separate(
    col = ninBeskrivelsesvariable,
    into = c("NiN_variable_code", "NiN_variable_value"),
    sep = "_",
    remove = F
  ) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
  select(
    id = identifikasjon_lokalId,
    municipality = kommunenummer,
    year = kartleggingsår,
    mosaic = mosaikk,
    quality = lokalitetskvalitet,
    biodiversity = naturmangfold,
    condition = tilstand,
    natureType = naturtype,
    variable = NiN_variable_code,
    value = NiN_variable_value,
    area
  ) |>
  st_transform(myCRS) # Chosing this to match the EDM (se further down)

# 13k obs.

@@

naturetypes <- naturetypes %>%
  mutate(value = case_when(
    # selecting the variable that follow the same 4 step scale
    variable %in% c("7TK", "7SE", "7FA") ~ 
      case_match(value,
             0 ~ 0,
             1 ~ mean(c(0, 1/16))*100,
             2 ~ 1/16*100,
             3 ~ 50), # note that it is not possible to get a value of 1
    # selecting the eight step variables
    variable %in% c('PRTK', 'PRSL') ~ 
      case_match(value,
             0 ~ 0,
             1 ~ 1.5,
             2 ~ 3,
             3 ~ 6.25,
             4 ~ 12.5,
             5 ~ 25,
             6 ~ 50,
             7 ~ 75)
    )
  )

@@

naturetypes_wide <- naturetypes |>
  # Column names starting with a number is problematic, so adding a prefix
  mutate(variable = paste0("var_", variable)) |>
  pivot_wider(
    names_from = "variable",
    values_from = "value",
    id_cols = "id") |>
    as_tibble()

@@

# Taking he sum of 7SE and 7TK (incl the PR.. variabels)
naturetypes_wide <- naturetypes_wide |>
  rowwise() |>
  mutate(ADSV = sum(c(SE, TK), na.rm=TRUE))

@@

```{r}
#| eval: true
# Now I will copy these ADVS-values into the sf object again
naturetypes <- naturetypes |>
  left_join(naturetypes_wide |> select(id, var_7FA, ADSV), by = "id")
```

@@

# Now I rescale the now continous variables using reference and threshold values
# I will use the same reference levels/values for all of Norway:

@@

@@

# same for 7FK
eaTools::ea_normalise(data = naturetypes,
                      vector = "var_7FA",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      plot=T,
                      reverse = T
                      )
# The variables are really coarse
```

@@

naturetypes$i_alien <- eaTools::ea_normalise(data = naturetypes,
                      vector = "var_7FA",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      reverse = T
                      )
```

@@

corrCheck |>
  as_tibble() |>
  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
         i_alien_fct = floor(round(i_alien*10,2))/10) |>
  count(i_ADSV_fct, infrastructureIndex) |>
  View()

# A first look
corrCheck |>
  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
         i_alien_fct = floor(round(i_alien*10,2))/10) |>
  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct),
               values_to = "indicatorValue",
               names_to = "indicator") |>
  filter(!is.na(indicatorValue)) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = factor(round(indicatorValue,2))))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_brewer(palette = "RdYlGn")+
  facet_grid(indicator~year)
```

@@

# Using a colour gradient emphasises the first colour (dark green).
# Lets try discrete colours, and merge som classes to simplify
corrCheck |>
  pivot_longer(cols = c(i_ADSV, i_alien),
               values_to = "indicatorValue",
               names_to = "indicator") |>
  filter(!is.na(indicatorValue)) |>
  mutate(condition = case_when(
    indicatorValue <0.6 ~"Poor",
    indicatorValue <0.8 ~"Intermediate",
    .default = "Good"
  )) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = condition))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_y_log10() +
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)

>>>>

```{r getRelevantNTs}
#| eval: true

@@

myVars <- c("7TK", "7SE", "PRTK", "PRSL", "7FA", "7GR-GI")
nts <- naturetypes_summary %>%
  rowwise() %>%
  mutate(keepers = sum(c_across(
    all_of(myVars))>0, na.rm=T)) |>
  filter(keepers >0,
         Ecosystem == "våtmark") |>
  pull(Nature_type)
```

@@

# Impart polygon data set
naturetypes <- sf::st_read(dsn = path_naturetypes)
# 117k polygons (2022)
# sometimes this function struggles with Norwegian letters.
# Restart R and try again.

naturetypes <- naturetypes |>
  # keep only wetlands
  filter(
    hovedøkosystem == "våtmark",
    naturtype %in% nts,
    naturtype != "Kalkrik helofyttsump"
  ) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep = ",") |>
  separate(
    col = ninBeskrivelsesvariable,
    into = c("NiN_variable_code", "NiN_variable_value"),
    sep = "_",
    remove = F
  ) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% myVars) |>
  select(
    id = identifikasjon_lokalId,
    municipality = kommunenummer,
    year = kartleggingsår,
    mosaic = mosaikk,
    quality = lokalitetskvalitet,
    biodiversity = naturmangfold,
    condition = tilstand,
    natureType = naturtype,
    variable = NiN_variable_code,
    value = NiN_variable_value,
    area
  ) |>
  st_transform(myCRS) # Choosing this to match the EDM (se further down)
# 19k obs.

naturetypes |>
  as_tibble() |>
  count(natureType, sort=T) |>
  mutate(natureType = fct_reorder(natureType, n)) |>
  ggplot(aes(x = natureType, y = n))+
  geom_col()+
  coord_flip()

@@

naturetypes <- naturetypes %>%
  mutate(value = case_when(
    # selecting the variable that follow the same 4 step scale
    variable %in% c("7TK", "7SE", "7FA") ~ 
      case_match(value,
             0 ~ 0,
             1 ~ mean(c(0, 1/16))*100,
             2 ~ 1/16*100,
             3 ~ 50), # note that it is not possible to get a value of 1
    # selecting the eight step variables
    variable %in% c('PRTK', 'PRSL') ~ 
      case_match(value,
             0 ~ 0,
             1 ~ 1.5,
             2 ~ 3,
             3 ~ 6.25,
             4 ~ 12.5,
             5 ~ 25,
             6 ~ 50,
             7 ~ 75),
    .default = value
    )
  )

@@

naturetypes_wide <- naturetypes |>
  filter(variable %in% c("7TK", "7SE", "PRTK", "PRSL")) |>
  # Column names starting with a number is problematic, so adding a prefix
  mutate(variable = paste0("var_", variable)) |>
  pivot_wider(
    names_from = "variable",
    values_from = "value",
    id_cols = "id") |>
    as_tibble()

@@

# Taking the sum of 7SE and 7TK (incl the PR.. variables)
naturetypes_wide <- naturetypes_wide |>
  rowwise() |>
  mutate(ADSV = sum(c(SE, TK), na.rm=TRUE))

@@

# Now I will copy these ADVS-values into the sf object again, keeping things in wide format
naturetypes <- naturetypes |>
  pivot_wider(#id_cols="id",
              names_from = "variable",
              values_from = "value") |>
  left_join(naturetypes_wide |> select(id, ADSV), by = "id") |>
  select(!c("7TK", "7SE", "PRSL", "PRTK")) 
  # >
  #pivot_longer(cols = c(ADSV, "7FA", "7GR-GI"),
  #             names_to = "variable",
  #             values_drop_na = T)
```

@@

# Now I rescale the now continuous variables using reference and threshold values
# I will use the same reference levels/values for all of Norway for ADSV and alien species:

@@

# For 7GR-GI I use this
upper2 <- 1
lower2 <- 5
threshold2 <- 2.5 # = observable effect. Value 3 indicates a shit to a new type (grunntype)

@@

# same for 7FK
eaTools::ea_normalise(data = naturetypes,
                      vector = "7FA",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      plot=T,
                      reverse = T
                      )
# The variables are really coarse

eaTools::ea_normalise(data = naturetypes,
                      vector = "7GR-GI",
                      upper_reference_level = lower2,
                      lower_reference_level = upper2,
                      break_point = threshold2,
                      plot=T,
                      reverse = T
                      )

@@

naturetypes$i_alien <- eaTools::ea_normalise(data = naturetypes,
                      vector = "7FA",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      reverse = T
                      )

naturetypes$i_ditch <- eaTools::ea_normalise(data = naturetypes,
                      vector = "7GR-GI",
                      upper_reference_level = lower2,
                      lower_reference_level = upper2,
                      break_point = threshold2,
                      reverse = T
                      )
```

@@

#corrCheck |>
#  as_tibble() |>
#  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
#         i_alien_fct = floor(round(i_alien*10,2))/10) |>
#  count(i_ADSV_fct, infrastructureIndex) |>
#  View()

# A first look
corrCheck |>
  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
         i_alien_fct = floor(round(i_alien*10,2))/10,
         i_ditch_fct = floor(round(i_ditch*10,2))/10) |>
  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na =T) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = factor(indicatorValue)))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_brewer(palette = "RdYlGn")+
  facet_grid(indicator~year)
```
```{r}
#| eval: false

# The year effect is constant, so ignoring that
corrCheck |>
  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
         i_alien_fct = floor(round(i_alien*10,2))/10,
         i_ditch_fct = floor(round(i_ditch*10,2))/10) |>
  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na =T) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = factor(indicatorValue)))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_brewer(palette = "RdYlGn")+
  facet_grid(~indicator)
```

@@

# Using a colour gradient emphasises the first colour (dark green).
# Lets try discrete colours, and merge som classes to simplify
corrCheck |>
  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na = T) |>
  mutate(condition = case_when(
    indicatorValue <0.6 ~"Poor",
    indicatorValue <0.8 ~"Intermediate",
    indicatorValue <0.91 ~"Good",
    .default = "Perfect"
  ),
  condition = fct_reorder(condition, indicatorValue)
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = condition))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)

```
```{r}
#| eval: false

# The figure changes a lot depending on which value gets the darker colour
corrCheck |>
  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na = T) |>
  mutate(condition = case_when(
    indicatorValue <0.6 ~"Poor",
    indicatorValue <0.8 ~"Intermediate",
    indicatorValue <0.91 ~"Good",
    .default = "Perfect"
  ),
  condition = fct_reorder(condition, -indicatorValue)
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = condition))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)
```

# If I look at this on a log scale, it also changes and becomes less convincing for ADSV and alien
corrCheck |>
  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na = T) |>
  mutate(condition = case_when(
    indicatorValue <0.6 ~"Poor",
    indicatorValue <0.8 ~"Intermediate",
    indicatorValue <0.91 ~"Good",
    .default = "Perfect"
  ),
  condition = fct_reorder(condition, -indicatorValue)
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = condition))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_y_log10()+
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)

# This is not correct I think as it evens out differences that are clearly there.
# i_ditch is showin less of a pattern for example
```

# HIAs 1 and 2 are quite similar, so perhaps I can combine them
corrCheck |>
  pivot_longer(cols = c(i_ADSV, i_alien),
               values_to = "indicatorValue",
               names_to = "indicator") |>
  filter(!is.na(indicatorValue)) |>
  mutate(condition = case_when(
    indicatorValue <0.6 ~"Poor",
    indicatorValue <0.8 ~"Intermediate",
    indicatorValue <0.91 ~"Good",
    .default = "Perfect"
  ),
  condition = fct_reorder(condition, indicatorValue),
  HIA = case_when(
    infrastructureIndex %in% c(1,2) ~ "1 & 2",
    .default = factor(infrastructureIndex)
  )
  ) |>
  ggplot(aes(
    x = HIA, 
    fill = condition))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_y_log10() +
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)
```
